The Concurrent Calculi Formalisation Benchmark

### 2024-02-21 short-paper/short-paper.tex
revisions to scope extr: still the "quotation" stuff to be fixed

\subsection{Scope extrusion}
This challenge revolves around the mechanisation of explicit scope
extrusion, by which a process can send restricted names to another
process, as long as the restriction can safely be extruded to include
the receiving process.  The setting for this challenge is a
``classic'' untyped \( \pi \)-calculus, where (unlike the calculi in
the other challenges) names can be sent and received, and bound by
input constructs.  We define two different semantics for our system:
\begin{enumerate}
\item A reduction system: this avoids explicit reasoning
about scope extrusion via structural congruence, allowing implementors to explore
alternative ways to encode the latter (\eg via context or via congruent refinements)
\item A (early) labelled transition system
\end{enumerate}
The objective of this
challenge is to prove that the two semantics are equivalent up to
structural congruence.

This is the challenge most closely related to the POPLMark challenge
since it concerns the properties of binders, whose encoding has been
extensively studied w.r.t.\ the $\lambda$-calculi. Process calculi
present additional challenges: $\pi$-calculi typically include several
different binding constructs: inputs bind a received name or value,
recursive processes bind recursion variables, and restrictions bind
names.  The first two act similarly to the binders known from
$\lambda$-calculi, but restrictions may be more challenging due to
scope extrusion.

Given those peculiarities, the syntax and semantics has been
mechanised from an early age (see~\cite{Melham1994}) with many proof
assistants and so many encoding styles that it would take too long to
review.  They range from concrete encodings based on named syntax
(Melham's op.cit.) to basic de Bruijn (cite) and locally nameless
representation~\cite{Chargueraud2012} (\eg~\cite{Castro2020}).
Nominal approaches are also common, see~\cite{Bengtson2009}, but they
may be problematic w.r.t.~proof assistants based on dependent types.

@@

Higher-order abstract syntax (HOAS) has seen extensive and successful
use in formal reasoning~\cite{Despeyroux2000,Honsell2001,Tiu2010}. Its
weak form aligns well with mainstream inductive proof assistants,
simplifying the encoding of typing systems and operational
semantics. However, when addressing more intricate concepts like
bisimulation, HOAS necessitates extensions. These extensions may take
the form of additional axioms~\cite{Honsell2001} or require specialised proof assistants
like Abella, which introduces a dedicated quantifier for handling
properties related to names~\cite{GacekMN11}.

@@

\subsubsection{Evaluation criteria.}
The idea behind our benchmark is to obtain evidence towards answering
questions \cref{item:rq1,item:rq2,item:rq3}. We are therefore
interested not only in the solutions, but also in the experience of
solving the challenges with the chosen approach.  Solutions to our
challenges should be compared on three axes:
\begin{enumerate}
\item Mechanisation overhead: the amount of manually written infrastructure and setup needed to express the definitions in the mechanisation;
\item Adequacy of the formal statements in the mechanisation: whether the proven theorems are easily recognisable as the theorems from the challenge; and
\item Cost of entry for the tools and techniques employed: the difficulty of learning to use the techniques.
\end{enumerate}
Solutions to our challenges do not need to strictly follow the definitions and lemmata set out in the problem text, but solutions which deviate from the original challenges must present more elaborate argumentation for their adequacy.

>>>>

\subsection{Scope extrusion}
This challenge revolves around the mechanisation of % explicit
scope
extrusion, by which a process can send restricted names to another
process, as long as the restriction can safely be extruded to include
the receiving process.  The setting for this challenge is a
``classic'' untyped \( \pi \)-calculus, where (unlike the calculi in
the other challenges) names can be sent and received, and bound by
input constructs.  We define two different semantics for our system:
\begin{enumerate}
\item A reduction system: this avoids explicit reasoning about scope
  extrusion by using instead structural congruence, allowing
  implementors to explore alternative ways to encode the latter (\eg
  via process contexts or via compatible refinement);
\item A (early) labelled transition system.
\end{enumerate}
The objective of this
challenge is to prove that the two semantics are equivalent up to
structural congruence.

This is the challenge most closely related to  POPLMark, 
since it concerns the properties of binders, whose encoding has been
extensively studied w.r.t.\ the $\lambda$-calculi. Process calculi
present additional challenges,  typically including several
different binding constructs: inputs bind a received name or value,
recursive processes bind recursion variables, and restrictions bind
names.  The first two act similarly to the binders known from
$\lambda$-calculi, but restrictions may be more challenging due to
scope extrusion.

Given those peculiarities, the syntax and semantics of $\pi$-calculi have been
mechanised from an early age (see~\cite{Melham1994}) with many proof
assistants and in so many encoding styles that it would take too long to
review.  They range from concrete encodings based on named syntax
(Melham's op.cit.) to basic de Bruijn (cite) and locally nameless
representation%~\cite{Chargueraud2012}
(\eg~\cite{Castro2020}).
Nominal approaches are also common, see~\cite{Bengtson2009}, but they
may be problematic w.r.t.~proof assistants based on dependent types.

@@

Higher-order abstract syntax (HOAS) has seen extensive and successful
use in formal reasoning in this area~\cite{Despeyroux2000,Honsell2001,Tiu2010}. Its
weak form aligns well with mainstream inductive proof assistants,
significantly simplifying the encoding of typing systems and operational
semantics. However, when addressing more intricate concepts like
bisimulation, HOAS necessitates extensions. These extensions may take
the form of additional axioms~\cite{Honsell2001} or require specialised proof assistants
like Abella, which introduces a dedicated quantifier for handling
properties related to names~\cite{GacekMN11}.

@@

\subsection{Evaluation criteria.}
The idea behind our benchmark is to obtain evidence towards answering
questions \cref{item:rq1,item:rq2,item:rq3}. We are therefore
interested not only in the solutions, but also in the experience of
solving the challenges with the chosen approach.  Solutions to our
challenges should be compared on three axes:
\begin{enumerate}
\item Mechanisation overhead: the amount of manually written infrastructure and setup needed to express the definitions in the mechanisation;
\item Adequacy of the formal statements in the mechanisation: whether the proven theorems are easily recognisable as the theorems from the challenge; and
\item Cost of entry for the tools and techniques employed: the difficulty of learning to use the techniques.
\end{enumerate}
Solutions to our challenges do not need to strictly follow the definitions and lemmata set out in the problem text, but solutions which deviate from the original challenges must present more elaborate argumentation for their adequacy.
